---
title: "cortisol practice polynomial and linear fits"
author: "gs"
date: "2023-09-19"
output: pdf_document
---


```{r}
library(psych)
library(ggplot2)
library(lme4)
#library(nlme)
library(readxl)
library(excel.link)
library(tidyverse)

cortisol_raw<-xl.read.file("C:/SalivaResults_10.31.2022_GS_fixednom.xlsx",
                           password="FIP2022")
comp_cort_raw<-xl.read.file("C:/SalivaResults_10.31.2022_GS_comp.xlsx",
                           password="FIP2022")

#Separating into dataframes for parent and child together and individually
comp_cort <- comp_cort_raw%>%
  select(cort_1, cort_2, cort_3, cort_4, cort_5, rID, part, C_P)

c_cort_wide <- na.omit(cortisol_raw) %>%
  select(rID, cortC_1, cortC_2, cortC_3, cortC_4, cortC_5, ERC_tot, ERC_hilo, ERQ, ERQ_s, ERQ_r)
p_cort_wide <- na.omit(cortisol_raw) %>%
  select(rID, cortP_1, cortP_2, cortP_3, cortP_4, cortP_5, ERC_tot, ERC_hilo, ERQ, ERQ_s, ERQ_r)

```

transforming to long format and scaling for time
```{r}
comp_cort_long <- reshape(data=na.omit(comp_cort), 
                         timevar=c("time"), 
                         idvar="rID",
                         varying=c("cort_1", "cort_2", "cort_3", "cort_4", "cort_5"),
                         direction="long", sep="_")

comp_cort_long$timescaled <- (comp_cort_long$time - 1)/5
```

```{r}
p_cort_long <- reshape(data=p_cort_wide, 
                         timevar=c("time"), 
                         idvar="rID",
                         varying=c("cortP_1", "cortP_2", "cortP_3", "cortP_4", "cortP_5"),
                         direction="long", sep="_")

p_cort_long$timescaled <- (p_cort_long$time - 1)/5

```

```{r}
c_cort_long <- reshape(data=c_cort_wide, 
                         timevar=c("time"), 
                         idvar="id",
                         varying=c("cortC_1", "cortC_2", "cortC_3", "cortC_4", "cortC_5"),
                         direction="long", sep="_")

c_cort_long$timescaled <- (c_cort_long$time - 1)/5
```


Plotting to see spread of data before fitting models
```{r eval = FALSE}
ggplot(data=na.omit(c_cort_long), aes(x=factor(time), y=cortC)) + 
  geom_boxplot(notch = FALSE) +
  stat_summary(fun="mean", geom="point", shape=23, size=3, fill="white") +
  labs(x = "Time", y = "Child Cortisol")
```
```{r}
ggplot(data=na.omit(p_cort_long), aes(x=factor(time), y=cortP)) + 
  geom_boxplot(notch = FALSE) +
  stat_summary(fun="mean", geom="point", shape=23, size=3, fill="white") +
  labs(x = "Time", y = "Parent Cortisol")
```


all data with color for parent or child
```{r eval = FALSE}
ggplot(data=na.omit(comp_cort_long), aes(x=factor(time), y=cort, color=C_P)) + 
   geom_boxplot(notch = FALSE) +
   stat_summary(fun="mean", geom="point", shape=23, size=3, fill="white") +
   labs(x = "Time", y = "Cortisol")

```

child data with ERC_hilo
```{r eval = FALSE}
ggplot(data=na.omit(c_cort_long), aes(x=factor(time), y=cortC, color=ERC_hilo)) +
   geom_boxplot(notch = FALSE) +
   stat_summary(fun="mean", geom="point", shape=23, size=3, fill="white") +
   labs(x = "Time", y = "Cortisol")
```

parent cortisol, child ERC
```{r eval = FALSE}
ggplot(data=na.omit(p_cort_long), aes(x=factor(time), y=cortP, color=ERC_hilo)) +
   geom_boxplot(notch = FALSE) +
   stat_summary(fun="mean", geom="point", shape=23, size=3, fill="white") +
   labs(x = "Time", y = "Cortisol")
```

```{r eval =FALSE}
ggplot(data=na.omit(p_cort_long), aes(x=factor(time), y=cortP, color=ERQ_s)) +
   geom_boxplot(notch = FALSE) +
   stat_summary(fun="mean", geom="point", shape=23, size=3, fill="white") +
   labs(x = "Time", y = "Cortisol")
```


linear best fit for parent cortisol with group effect ERC
```{r}
p_cort_m_base <-lmer(cortP~1+time+(1+time|rID),data=p_cort_long, REML=FALSE)
p_cort_m0 <- lmer(cortP~1+time+ERC_hilo+(1+time|rID),data=p_cort_long, REML=FALSE)
p_cort_m1 <- lmer(cortP~time*ERC_hilo+(time|rID),data=p_cort_long, REML=FALSE)

anova(p_cort_m_base, p_cort_m1, p_cort_m0)
```

linear for parent cortisol with group effects for ERQ_s
```{r}
p_cort_ERQ_s_m_base <-lmer(cortP~1+time+(1+time|rID),data=p_cort_long, REML=FALSE)
p_cort_ERQ_s_m0 <- lmer(cortP~1+time+ERQ_s+(1+time|rID),data=p_cort_long, REML=FALSE)
p_cort_ERQ_s_m1 <- lmer(cortP~time*ERQ_s+(time|rID),data=p_cort_long, REML=FALSE)

anova(p_cort_ERQ_s_m_base, p_cort_ERQ_s_m1, p_cort_ERQ_s_m0)
```

plotting the models from above
```{r}
ggplot(p_cort_long, aes(time, cortP, color=ERQ_s))+stat_summary(fun.data=mean_se, geom="pointrange")+ stat_summary(aes(y=fitted(p_cort_ERQ_s_m0),linetype=ERQ_s), fun=mean, geom="line")+
  labs(x = "Time", y = "Cortisol")
```

Growth curve modeling (via Mirman, D. (2014). Growth Curve Analysis and Visualization Using R. Chapman and Hall / CRC.)
```{r}
p_t = poly(unique(p_cort_long$time),2)

p_cort_long[,paste("p_ot", 1:2, sep="")] <- p_t[p_cort_long$time,1:2]

cort_p_ot_m <-lmer(cortP~(p_ot1+p_ot2)+(p_ot1+p_ot2|rID),data=p_cort_long, REML=FALSE)
#no effect for ERQ_s

cort_p_ot_m0 <-lmer(cortP~(p_ot1+p_ot2)+ERQ_s+(p_ot1+p_ot2|rID),data=p_cort_long, REML=FALSE)
#fixed effect of ERQ_s on intercept

cort_p_ot_m1 <-lmer(cortP~(p_ot1+p_ot2)+ERQ_s+p_ot1:ERQ_s+(p_ot1+p_ot2|rID),data=p_cort_long, REML=FALSE)
#effect of ERQ_s on linear term only

cort_p_ot_m2 <-lmer(cortP~(p_ot1+p_ot2)*ERQ_s+(p_ot1+p_ot2|rID),data=p_cort_long, REML=FALSE)
#effect of ERQ_s on all time terms
anova(cort_p_ot_m, cort_p_ot_m0, cort_p_ot_m1, cort_p_ot_m2)
```

plotting fitted models...
```{r}
ggplot(na.omit(p_cort_long), aes(time, cortP, color=ERQ_s))+ 
  stat_summary(aes(y=fitted(cort_p_ot_m1),linetype=ERQ_s), fun=mean, geom="line", linewidth=.75)+
  stat_summary(fun.data=mean_se, geom="pointrange", linewidth=.75)+
  theme_bw(base_size=10)+
  #coord_cartesian(ylim=c(0.5,1.0))+
  #scale_x_continuous(breaks=1.00)+
  labs(x = "Time", y = "Parent Cortisol", title = "Parent Cortisol across ERQ Supression Subscale")
```

linear best fit for child with group effect ERC
```{r}

c_cort_m_base <-lmer(cortC~1+time+(1+time|rID),data=c_cort_long, REML=FALSE)
c_cort_m0 <- lmer(cortC~1+time+ERC_hilo+(1+time|rID),data=c_cort_long, REML=FALSE)
c_cort_m1 <- lmer(cortC~time*ERC_hilo+(time|rID),data=c_cort_long, REML=FALSE)

anova(c_cort_m_base, c_cort_m1, c_cort_m0)
```
m1 is best but not great

plot of m1 for ERC effects
```{r eval = FALSE}
ggplot(c_cort_long, aes(time, cortC, color=ERC_hilo))+stat_summary(fun.data=mean_se, geom="pointrange")+ stat_summary(aes(y=fitted(c_cort_m1),linetype=ERC_hilo), fun=mean, geom="line")+
  labs(x = "Time", y = "Cortisol")
```

polynomial best fit for child with ERC
```{r}
c_t = poly(unique(c_cort_long$time),2)

c_cort_long[,paste("c_ot", 1:2, sep="")] <- c_t[c_cort_long$time,1:2]

cort_c_ot_m <-lmer(cortC~(c_ot1+c_ot2)+(c_ot1+c_ot2|rID),data=c_cort_long, REML=FALSE)
#no effect for ERC_hilo

cort_c_ot_m0 <-lmer(cortC~(c_ot1+c_ot2)+ERC_hilo+(c_ot1+c_ot2|rID),data=c_cort_long, REML=FALSE)
#fixed effect of ERC_hilo on intercept

cort_c_ot_m1 <-lmer(cortC~(c_ot1+c_ot2)+ERC_hilo+c_ot1:ERC_hilo+(c_ot1+c_ot2|rID),data=c_cort_long, REML=FALSE)
#effect of ERC_hilo on linear term only

cort_c_ot_m2 <-lmer(cortC~(c_ot1+c_ot2)*ERC_hilo+(c_ot1+c_ot2|rID),data=c_cort_long, REML=FALSE)
#effect of ERC_hilo on all time terms
anova(cort_c_ot_m, cort_c_ot_m0, cort_c_ot_m1, cort_c_ot_m2)
#anova(cort_m_base, cort_m1, cort_m0, cort_ot_m, cort_ot_m0, cort_ot_m1, cort_ot_m2)
```
m1 still best but not great


plotting new m1
```{r eval = FALSE}
ggplot(na.omit(c_cort_long), aes(time, cortC, color=ERC_hilo))+ 
  stat_summary(aes(y=fitted(cort_c_ot_m),linetype=ERC_hilo), fun=mean, geom="line", linewidth=.75)+
  stat_summary(fun.data=mean_se, geom="pointrange", linewidth=.75)+
  theme_bw(base_size=10)+
  #coord_cartesian(ylim=c(0.5,1.0))+
  #scale_x_continuous(breaks=1.00)+
  labs(x = "Time", y = "Cortisol")
```


linear best fit for all data with group effects for child and parent
```{r}
cort_m_base <-lmer(cort~1+time+(1+time|rID),data=comp_cort_long, REML=FALSE)
cort_m0 <- lmer(cort~1+time+C_P+(1+time|rID),data=comp_cort_long, REML=FALSE)
cort_m1 <- lmer(cort~time*C_P+(time|rID),data=comp_cort_long, REML=FALSE)

anova(cort_m_base, cort_m1, cort_m0)
  
```
m1 best 

```{r}
ggplot(comp_cort_long, aes(time, cort, color=C_P))+stat_summary(fun.data=mean_se, geom="pointrange")+ stat_summary(aes(y=fitted(cort_m1),linetype=C_P), fun=mean, geom="line")+
  labs(x = "Time", y = "Cortisol")
```

```{r}
t = poly(unique(comp_cort_long$time),2)

comp_cort_long[,paste("ot", 1:2, sep="")] <- t[comp_cort_long$time,1:2]

cort_ot_m <-lmer(cort~(ot1+ot2)+(ot1+ot2|rID),data=comp_cort_long, REML=FALSE)
#no effect for C_P

cort_ot_m0 <-lmer(cort~(ot1+ot2)+C_P+(ot1+ot2|rID),data=comp_cort_long, REML=FALSE)
#fixed effect of C_P on intercept

cort_ot_m1 <-lmer(cort~(ot1+ot2)+C_P+ot1:C_P+(ot1+ot2|rID),data=comp_cort_long, REML=FALSE)
#effect of C_P on linear term only

cort_ot_m2 <-lmer(cort~(ot1+ot2)*C_P+(ot1+ot2|rID),data=comp_cort_long, REML=FALSE)
#effect of C_P on all time terms
anova(cort_ot_m, cort_ot_m0, cort_ot_m1, cort_ot_m2)
#anova(cort_m_base, cort_m1, cort_m0, cort_ot_m, cort_ot_m0, cort_ot_m1, cort_ot_m2)



```
m1 best again... linear m1 may be better than polynomial m1?

now let's plot those polynomial values 
```{r}
ggplot(na.omit(comp_cort_long), aes(time, cort, color=C_P))+ 
  stat_summary(aes(y=fitted(cort_ot_m),linetype=C_P), fun=mean, geom="line", linewidth=.75)+
  stat_summary(fun.data=mean_se, geom="pointrange", linewidth=.75)+
  theme_bw(base_size=10)+
  #coord_cartesian(ylim=c(0.5,1.0))+
  #scale_x_continuous(breaks=1.00)+
  labs(x = "Time", y = "Cortisol")
```

parameter estimates
```{r}
#fixed effect parameter estimates and their p-values:
cort_coefs<-data.frame(coef(summary(cort_ot_m1)))
cort_coefs%>%
  mutate(p=2*(1-pnorm(abs(cort_coefs$t.value))))

#show parameter specific p-values:
library(lmerTest)
cort_ot_m1t<-lmer(cort~(ot1+ot2)+C_P+ot1:C_P+(ot1+ot2|rID),data=comp_cort_long, REML=FALSE)
coef(summary(cort_ot_m1t))
```






```{r}
round(cor(p_cort_wide[,c("cortP_1", "cortP_2", "cortP_3", "cortP_4", "cortP_5")], use="complete.obs",method="spearman"),2)
```

```{r}
round(cor(c_cort_wide[,c("cortC_1", "cortC_2", "cortC_3", "cortC_4", "cortC_5")], use="complete.obs",method="spearman"),2)
```

```{r eval = FALSE}
round(cor(cortisol_raw[,c("cortC_1", "cortC_2", "cortC_3", "cortC_4", "cortC_5", "cortP_1", "cortP_2", "cortP_3", "cortP_4", "cortP_5")], use="complete.obs",method="spearman"),2)
```

```{r}
ggplot(data = c_cort_long, aes(x = time, y = cortC, group = id)) +
  geom_point(color="black") + 
  geom_line(color="black") +
  xlab("Time") + 
  ylab("Cortisol") + ylim(0,.5) +
  scale_x_continuous(breaks=seq(0,8,by=1)) 
```

```{r eval = FALSE}
library(nlme)
c_cort_linear <- nlme(cortC ~ g0 + g1*timescaled,
                    fixed = g0 + g1 ~ 1,
                    random = g0 + g1 ~ 1,
                    group = ~id,
                    start = c(g0=.05, g1=.3),
                    data = c_cort_long,
                    na.action = "na.exclude")

summary(c_cort_linear)  
```


```{r eval = FALSE}
p_cort_linear <- nlme(cortP ~ g0 + g1*timescaled,
                    fixed = g0 + g1 ~ 1,
                    random = g0 + g1 ~ 1,
                    group = ~rID,
                    start = c(g0=.05, g1=.3),
                    data = p_cort_long,
                    na.action = "na.exclude")

summary(c_cort_linear)  
```

```{r eval = FALSE}
c_cort_long$pred_linear <- predict(c_cort_linear)

#obtaining predicted scores for prototype
c_cort_long$proto_linear <- predict(c_cort_linear, level=0)

#plotting predicted trajectories
#intraindividual change trajetories
ggplot(data = c_cort_long, aes(x = time, y = pred_linear, group = rID)) +
  geom_line(color="black") +
  geom_line(aes(x = time, y = proto_linear), color="red",linewidth=2) + 
  xlab("Time") + 
  ylab("Cortisol") + ylim(0,0.5) +
  scale_x_continuous(breaks=seq(1,5,by=1)) 
```
```{r eval = FALSE}
p_cort_long$pred_linear <- predict(p_cort_linear)

#obtaining predicted scores for prototype
p_cort_long$proto_linear <- predict(p_cort_linear, level=0)

#plotting predicted trajectories
#intraindividual change trajetories
ggplot(data = p_cort_long, aes(x = time, y = pred_linear, group = rID)) +
  geom_line(color="black") +
  geom_line(aes(x = time, y = proto_linear), color="red",linewidth=2) + 
  xlab("Time") + 
  ylab("Cortisol") + ylim(0,0.5) +
  scale_x_continuous(breaks=seq(1,5,by=1))
```
summary of linear mixed effects for both parents and children
```{r eval = FALSE}
comp_cort_long$timesq <- comp_cort_long$timescaled^2
comp_cort_quad <- lmer(cort ~ 1 + timescaled + timesq + (1 + timescaled + timesq | rID),
                    data = comp_cort_long,
                    )
summary(comp_cort_quad)


# p_cort_long$timesq <- p_cort_long$timescaled^2
# p_cort_quad <- lmer(cortP ~ 1 + timescaled + timesq + (1 + timescaled + timesq | rID),
#                     data = p_cort_long,
#                     )
# summary(p_cort_quad)
```


```{r eval = FALSE}
library(nlme)
comp_cort_quad <- nlme(cort ~ g0 + g1*timescaled + g2*timesq,
                    fixed = g0 + g1 + g2 ~ 1,
                    random = g0 + g1 + g2 ~ 1,
                    group = ~rID,
                    start = c(g0=.5, g1=41.1, g2=-30.2),
                    data = comp_cort_long,
                    na.action = "na.exclude")

summary(comp_cort_quad) 
```

```{r eval = FALSE}
comp_cort_long$pred_quad <- predict(comp_cort_quad)

#obtaining predicted scores for prototype
comp_cort_long$proto_quad <- predict(comp_cort_quad, level=0)

#plotting predicted trajectories
#intraindividual change trajetories
# ggplot(data = comp_cort_long, aes(x = time, y = pred_quad, group = rID)) +
#   #geom_point(color="black") + 
#   geom_line(color="black") +
#   geom_line(aes(x = time, y = proto_quad), color="red",size=1.5) + 
#   xlab("Time") + 
#   ylab("Cortisol") + ylim(0,.6) +
#   scale_x_continuous(breaks=seq(1,5,by=1)) 

ggplot(na.omit(comp_cort_long), aes(time, cort, color=C_P))+ 
  stat_summary(aes(y=pred_quad,linetype=C_P), fun=mean, geom="line", linewidth=.75)+
  stat_summary(fun.data=mean_se, geom="pointrange", linewidth=.75)+
  theme_bw(base_size=10)+
  #coord_cartesian(ylim=c(0.5,1.0))+
  #scale_x_continuous(breaks=1.00)+
  labs(x = "Time", y = "Cortisol")
```

```{r eval = FALSE}
p_cort_quad <- nlme(cortP ~ g0 + g1*timescaled + g2*timesq,
                    fixed = g0 + g1 + g2 ~ 1,
                    random = g0 + g1 + g2 ~ 1,
                    group = ~rID,
                    start = c(g0=.5, g1=41.1, g2=-30.2),
                    data = p_cort_long,
                    na.action = "na.exclude")

#summary(p_cort_quad) 
p_cort_long$pred_quad <- predict(p_cort_quad)

#obtaining predicted scores for prototype
p_cort_long$proto_quad <- predict(p_cort_quad, level=0)

#plotting predicted trajectories
#intraindividual change trajetories
ggplot(data = p_cort_long, aes(x = time, y = pred_quad, group = rID)) +
  #geom_point(color="black") + 
  geom_line(color="black") +
  geom_line(aes(x = time, y = proto_quad), color="red",size=1) + 
  xlab("Time") + 
  ylab("Parent Cortisol") + ylim(0,.5) +
  scale_x_continuous(breaks=seq(1,5,by=1)) 
```


